% Summer Bridge R, Module 7: Analysis
% August 2019, UC Berkeley
% Chris Paciorek

```{r chunksetup, include=FALSE} 
# include any code here you don't want to show up in the document,
# e.g. package and dataset loading
library(fields)
library(foreign)
if(!('modules' %in% unlist(strsplit(getwd(), split = '/')))) setwd('modules')
gap <- read.csv(file.path('..', 'data', 'gapminder-FiveYearData.csv'), stringsAsFactors = FALSE)
gap2007 <- gap[gap$year == 2007, ]
```

# Describing relationships

- Once we've carried out group-wise operations and perhaps reshaped it, we may also like to describe the relationships in the data. Often this involves fitting some style of regression model.  The goal can be pure prediction, description, or inferring a causal relationship between variables.

Of course to infer causality, one has to be quite careful and techniques that try to avoid the usual pitfall that correlation is not causation are way beyond what we can cover here.

We'll just see the basics of how to fit regressions here. 

# Inference/Regression

- Running regressions in R is generally straightforward.

- Most basic, catch-all regression function in R is *glm*

- *glm* fits a generalized linear model with your choice of family/link function (gaussian, logit, poisson, etc.)

- *lm* is just a standard linear regression (equivalent to glm with family = gaussian(link = "identity"))

- The basic glm call looks something like this:

```{r eval=FALSE}
glm(formula = y ~ x1 + x2 + x3 + ..., family = familyname(link = "linkname"),
            data = )
```

- There are a bunch of families and links to use (help(family) for a full list), but some essentials are **binomial(link = "logit")**, **gaussian(link = "identity")**, and **poisson(link = "log")**

If you're using `lm`, the call looks the same but without the `family` argument. 

- Example: suppose we want to regress the life expectency on the GDP per capita and the population, as well as the continent and year.  The lm/glm call would be something like this:

```{r}
reg <- lm(formula = lifeExp ~ log(gdpPercap) + log(pop) + continent + year, 
                data = gap)
```

# Regression output

- When we store this regression in an object, we get access to several items of interest

```{r}
# View components contained in the regression output
names(reg)
# Examine regression coefficients
reg$coefficients
# Examine regression degrees of freedom
reg$df.residual
# See the standard (diagnostic) plots for a regression
plot(reg)
```

- R has a helpful summary method for regression objects
```{r}
summary(reg)
```

- Can also extract useful things from the summary object

```{r}
# Store summary method results
summ_reg <- summary(reg)
# View summary method results objects
objects(summ_reg)
# View table of coefficients
summ_reg$coefficients
```

- Note that, in our results, R has broken up our variables into their different factor levels (as it will do whenever your regressors have factor levels)

- If your data aren't factorized, you can tell lm/glm to factorize a variable (i.e. create dummy variables on the fly) by writing

```{r, eval=FALSE}
glm(formula = y ~ x1 + x2 + factor(x3), family = family(link = "link"),
            data = )
```

# Setting up regression interactions

- There are also some useful shortcuts for regressing on interaction terms:

`x1:x2` interacts all terms in x1 with all terms in x2
```{r}
summary(lm(lifeExp ~ log(gdpPercap) + log(pop) +
                    continent:factor(year), 
                    data = gap))
```

`x1*x2` produces the cross of x1 and x2, or x1+x2+x1:x2
```{r}
summary(lm(lifeExp ~ log(gdpPercap) + log(pop) + continent*factor(year), 
                data = gap))
```


# Smoothing

Linear regression and GLMs are of course useful, but often the relationship is not linear, even on some transformed scale.

Additive models and generalized additive models (GAMs) are the more flexible variants on linear models and GLMs.

There are a variety of tools in R for modeling nonlinear and smooth relationships, mirroring the variety of methods in the literature.

One workhorse is `gam()` in the *mgcv* package.

# GAM in action

Do we think there should be a linear relationship of life expectancy with GDP and year?

```{r gamExample, cache=TRUE, fig.width=10, fig.cap = ""}
library(mgcv)

mod <- gam(lifeExp ~ s(gdpPercap, k = 30) + s(year, k = 10), data = gap)

plot(mod)
summary(mod)

mod2 <- gam(lifeExp ~ s(log(gdpPercap), k = 30) + s(year, k = 10), data = gap)
plot(mod2)
```

If we were serious about building a good-fitting model, we could use the same kind of functionality as in lm/glm in terms of factors and interactions.

# How does GAM choose how much to smooth?

GAM uses the data to choose how much smoothing to do. Roughly one can think of what it is doing as carrying out cross-validation and choosing the best amount of smoothing for predicting held-out data.

`k` simply sets an upper bound on the amount of smoothing (you can think if `k` as the number of degrees of freedom - one would be a linear fit).

 - Make sure `k` is less than the number of unique values of the predictor variable
 - The default for `k` is relatively small and in some cases this may overly limit the smoothness of the curve.
    - You can try increasing `k` and see if it increases the fit.
    - If `summary` reports and `edf` that is close to `k` that often suggests that `k` should be increased.


# Optimization

R provides functionality for optimization - finding maxima or minima of a function. 

A workhorse is `optim()`, which implements a number of optimization algorithms. 

```{r eval=FALSE, fig.cap = ""} 
library(fields)  
```

```{r fig.width=11, fig.height=5, fig.cap = ""}
 banana <- function(x) {   ## Rosenbrock Banana function
         x1 <- x[1]
         x2 <- x[2]
         100 * (x2 - x1 * x1)^2 + (1 - x1)^2
     }

x1s <- x2s <- seq(-5, 5, length = 100)
x <- expand.grid(x1s, x2s)
fx <- apply(x, 1, banana)

par(mfrow = c(1, 2), mai = c(.45, .4, .1, .4))
image.plot(x1s, x2s, matrix(fx, 100), xlab = '', ylab = '')
image.plot(x1s, x2s, matrix(log(fx), 100), xlab = '', ylab = '')

optim(c(-2,0), banana)
```
We can see the progression of evaluations of the objective function:
```{r eval=FALSE, fig.cap = ""}
banana <- function(x) {   ## Rosenbrock Banana function
         points(x[1],x[2])
         Sys.sleep(.03)
         x1 <- x[1]
         x2 <- x[2]
         100 * (x2 - x1 * x1)^2 + (1 - x1)^2
     }
par(mfrow = c(1, 1), mai = c(.45, .4, .1, .4))
image.plot(x1s, x2s, matrix(log(fx), 100), xlab = '', ylab = '')
optim(c(-2,0), banana)
```



# Breakout 


### Using the ideas

1) Fit two linear regression models from the gapminder data, where the outcome is `lifeExp` and the explanatory variables are `log(pop)`, `log(gdpPercap)`, and `year`. In one model, treat `year` as a numeric variable. In the other, factorize the `year` variable. How do you interpret each model?

2) Modify the GAMs of lifeExp on gdpPercap and set `k` to a variety of values and see how the estimated relationships change. What about the estimated uncertainties?

### Advanced 

3) Fit a logistic regression model where the outcome is whether `lifeExp` is greater than or less than 60 years, exploring the use of different predictors.



